Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-16 main.Rmd
review till cyclic calendar algebra

<!-- Finally, visualizing data across single/multiple order-up granularities help us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. However, this approach can lead to a myriad of choices all of which are not useful. Analysts are expected to alliteratively visualize these choices for exploring possible patterns in the data. But too many choices might leave him bewildered.  -->

@@

\autoref{fig:circular-dow} depicts the linear (a and b) and the corresponding stack (c and d) display of time. The circular granularity day-of-week and day-of-fortnight are constructed from linear granularities day ($B$), week ($G$) and fortnight ($H$). The granules that overlap share elements from the underlying time domain in the linear display. The circular granularity $C_{B , G}$ and $C_{B, H}$ <!--= C_{\text{day}, \text{week}}$--> representing day-of-week and day-of-fortnight respectively consist of repeated pattern of categories $\{0, 1, 2, \dots, 6\}$ and $\{0, 1, 2, \dots, 13\}$ with $P=7$ and $P=14$ respectively. The granules that overlap represents the same category in the stack display. Each circular granularity can use descriptive label mappings, or be simply labelled using the index. <!--Similar to linear granularities, the indexing for each granule of the circular granularity can also be associated with textual representation, called the label.--> Suppose ${L}$ is a label mapping that defines an unique label for each index $l \in \{ 0,1,\dots, (P-1)\}$, then the label mapping $L$ for $C_{B, G}$ can be defined as
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sun}, \text{Mon}, \dots, \text{Sat}\}
$$
or
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}
$$
for example.

@@

A **quasi-circular** granularity can not be defined using modular arithmetic since they are formed using two linear granularities with irregular mapping. However, they are still formed with linear granularities, one of which "groups periodically into" the other. \autoref{tab:quasi} shows some example of quasi-circular granularities ($Q_i$) with ($P_i$) denoting the plausible choices of number of granules of the finer granularity inside each granule of the coarser one.

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
quasi-circular granularity && potential period lengths\\
\hline
$Q_1 =$ day-of-month && $P_1 = 31, 30, 29, 28$\\
$Q_2 =$ hour-of-month && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
$Q_3 =$ day-of-year && $P_3 = 366, 365$\\
$Q_4 =$ week-of-month && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of quasi-circular granularities $Q_i$ with potential period lengths $P_i$. These quasi-circular granularities relate two linear granularities one of which groups periodically into the other with irregular mapping leading many potential period lengths and hence can not be expressed through modular arithmetic. }
\label{tab:quasi}
\end{table}

\begin{definition}\label{def:quasicircular}
A {\bf quasi-circular granularity} $Q_{B, G'}$ relates linear granularities $G'$ and bottom granularity $B$, if
\begin{equation}\label{eq:quasi}
Q_{B, G'}(z) = z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert,for\quad z \in T_k
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$B$ denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G'$ with irregular mapping, $P'$ and $R'$ are the period of the grouping $(B, G')$ and the number of granules of $G'$ in $P'$, $T_w$ are the sets of indices of $B$ describing $G'(w)$ such that  $G'(w) = \bigcup_{z \in T_w}B(z)$ and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{definition}

@@

Consider a linear granularity $G'$ such that $G' = Alter^{2}_{(1,-2)}(B, Group_7(B))$, which implies it is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules.<!-- $$G'= Alter^{2}_{(1,-1)}(BG, Group_3(B))$$ and --><!--This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different.--> Then $Q_{B, G'}$ is a repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$. Here, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$. Hence using \ref{def:quasicircular}, then we will have:

@@

Consider a school semester which lasts for 18.3 weeks, starting with one week of orientation followed by an in-session period of 6 weeks, a semester-break of 1 week and again an in-session period of 7 weeks after which there is a study break of 1 week before final exams which continue for 1.3 weeks. Let the linear granularities $M_1$ and $M_2$ denote in-session semester period and semester break periods respectively. Both $M_1$ and $M_2$ and $M = M_{1}\bigcup M_{2}$ denoting the semester week type are aperiodic with respect to days ($B$) or weeks ($G$). Hence $A_{B, M}$ denoting day of the semester week type would be an aperiodic cyclic granularity. This is because the placement of the semester within an year would vary across years (\autoref{fig:aperiodic-example} c). \autoref{fig:aperiodic-example} b shows the stack display of time with granules representing same categories (semester break/in-session) stacked on top of one another. \autoref{fig:aperiodic-example} (a) representing the linear display of time is useful in manifesting how the categories would be determined with respect to the bottom granularities using Definition \ref{def:aperiodic}. It is interesting to note here that $Q_{H, M}$ denoting semester week of the semester week type would be a quasi-circular granularity since the distribution of semester weeks within a semester remains constant over years. Here $Q_{H, M}$ with a period of 18.3 weeks will have irregular mapping since each semester week types consists of different number of semester weeks.

@@

```{r aperiodic-example, echo=FALSE, out.width = "100%", fig.cap = "Quasi-circular and aperiodic cyclic granularities illustrated through linear (a) and stack (b) display of time. The linear display shows linear granularities days, weeks, semester weeks, semester week type distributed linearly from past to future. Here a semester lasts for 18.3 weeks, each starting with one week of orientation followed by an in-session period of 6 weeks, a semester-break of 1 week and again an in-session period of 7 weeks after which there is a study break of 1 week before final exams which continue for 1.3 weeks. This pattern remains same for all semester and hence $Q_{H, M}$ with $P = 18.3 weeks$ denoting week of the semester week type will be a quasi-circular granularity with repeating patterns. $A_{B, M}$ will be an aperiodic cyclic granularity since the placement of the semester within an year varies across years (shown through c)."}

@@

**Hierarchy table** Let $H_n: (G, C, K)$ be a hierarchy table containing $n$ linear granularities. $G_{l}$ and $G_{m}$ represent the linear granularity of order $l$ and $m$ respectively with $l<m$. $K \equiv P(l,m)$ represents the period length of the grouping $(G_{l}, G_{m})$ for circular granularity and $K \equiv k(l,m)$ represents the operation to obtain $G_{m}$ from $G_{l}$ for quasi-circular granularity $C_{G(l),G(m)}$, $\forall l,m \in {1, 2, \dots,n}$. We assume that the cyclic granularity $C_{G(l),G(m)}$ is either circular or quasi-circular. It is difficult to position an aperiodic linear granularity in a hierarchy and hence the possibility of an aperiodic cyclic granularity is not considered.

@@

<!-- The multiple order-up cyclic granularites resulting from the first two cases are circular and has been handled in \autoref{sec:all-circular}. Any cyclic granularity resulting from the last case are quasi-circular. Examples might include hour-of-month or day-of-year. -->

We consider the case of only one quasi-circular single order-up granularity in the hierarchy table. Any multiple order-up quasi-circular granularity $C_{l, m}(z)$ can then be obtained as a function of a single (or multiple) order-up circular granularity $C_{l,m'}(z)$ and single order-up quasi-circular granularity $C_{m',m}(z)$ or vice versa. Two different approaches need to be employed for the following cases:

<!-- For example, with $C_{l, m}(z) = f(C_{l,m'}(z), C_{m',m}(z))$, -->
- $C_{l,m'}(z)$ is circular and $C_{m',m}(z)$ is quasi-circular
\begin{equation} \label{eq:multifromsingle-quasi1}
\begin{split}
C_{G_l,G_{m}}(z) & = C_{G_{l},G_{m'}}(z) + P(l, m')(C_{G_{m'},G_{m}}(z)) \\
\end{split}
\end{equation}

@@

<!-- CONTENT CUT -->
<!-- Clearly, the computation of multiple order-up granularities become more involved with quasi-circular granularities. $C_{hour, month}(z)$ can be represented as function of circular granularity $C_{hour, day}(z)$ and quasi-circular granularity $C_{day, month}(z)$, both of which are single order-up. However, $C_{day, year}(z)$ can not be computed only with single order-up granularities $C_{day, month}(z)$ and $C_{month, year}(z)$. It also requires the knowledge of the composition of the linear granularity $days$ within $months$ in an $year$ for computation of $\vert T_w \vert$. -->
<!-- If $C_{G_{l},G_{m'}}(z)$ or $C_{G_{l},G_{m'}}(z)$ are multiple order-up circular granularities, they could be computed using similar approach in Section \ref{sec:all-circular}. -->

@@

<!-- CONTENT CHANGE -->
<!-- The representation of single order-up quasi-circular granularities using multiple order-up quasi-circular granularities is not discussed in this paper and is left for future work. -->

This case is similar to \ref{eq:multifromsingle-quasi2}. Single-order-up quasi-circular granularity ($C_{G_{l},G_{m'}}$) could be obtained from multiple-order-up quasi-circular granularity ($C_{G_l,G_{m}}$) and single/multiple order-up circular granularity $C_{m',m}$ using \ref{eq:multifromsingle-quasi2}.

>>>>

<!-- Finally, visualizing data across single/multiple-order-up granularities help us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. However, this approach can lead to a myriad of choices all of which are not useful. Analysts are expected to alliteratively visualize these choices for exploring possible patterns in the data. But too many choices might leave him bewildered.  -->

@@

\autoref{fig:circular-dow} depicts the linear (a and b) and the corresponding stack (c and d) display of time. Circular granularities day-of-week $C_{B , G}$ and day-of-fortnight $C_{B, H}$ are constructed from linear granularities day ($B$), week ($G$) and fortnight ($H$). The granules that overlap share elements from the underlying time domain in the linear display. The circular granularities $C_{B , G}$ and $C_{B , H}$  <!--= C_{\text{day}, \text{week}}$ representing day-of-week and day-of-fortnight respectively --> consist of repeated pattern of categories $\{0, 1, 2, \dots, 6\}$ and $\{0, 1, 2, \dots, 13\}$ with $P=7$ and $P=14$ respectively. The granules that overlap represents the same category in the stack display. Each circular granularity can use descriptive label mappings, or be simply labelled using the index. <!--Similar to linear granularities, the indexing for each granule of the circular granularity can also be associated with textual representation, called the label.--> Suppose ${L}$ is a label mapping that defines an unique label for each index $l \in \{ 0,1,\dots, (P-1)\}$, then the label mapping $L$ for $C_{B, G}$ can be defined as
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sun}, \text{Mon}, \dots, \text{Sat}\}
$$
or
$$
  L: \{0,1,2, \dots, 6\} \longmapsto\ \{\text{Sunday}, \text{Monday}, \dots, \text{Saturday}\}
$$
for example.

@@

A **quasi-circular** granularity can not be defined using modular arithmetic since they are formed using two linear granularities with irregular mapping. However, they are still formed with linear granularities, one of which "groups periodically into" the other. \autoref{tab:quasi} shows some example of quasi-circular granularities ($Q_i$) with ($P_i$) denoting the plausible choices of period of the grouping of two linear granularities. 
<!-- number of granules of the finer granularity inside each granule of the coarser one. -->

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
quasi-circular granularity && potential period lengths\\
\hline
$Q_1 =$ day-of-month && $P_1 = 31, 30, 29, 28$\\
$Q_2 =$ hour-of-month && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
$Q_3 =$ day-of-year && $P_3 = 366, 365$\\
$Q_4 =$ week-of-month && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of quasi-circular granularities $Q_i$ with potential period lengths $P_i$. These quasi-circular granularities relate two linear granularities one of which groups periodically into the other with irregular mapping leading to many potential period lengths and hence can not be expressed through modular arithmetic. }
\label{tab:quasi}
\end{table}

\begin{definition}\label{def:quasicircular}
A {\bf quasi-circular granularity} $Q_{B, G'}$ relates linear granularities $G'$ and bottom granularity $B$, if
\begin{equation}\label{eq:quasi}
\begin{split}
Q_{B, G'}(z) =
z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert, & \quad \text{for}\quad z \in T_{k} 
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$B$ denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G'$ with irregular mapping, $P'$ and $R'$ are the period of the grouping $(B, G')$ and the number of granules of $G'$ in $P'$, $T_w$ are the sets of indices of $B$ describing $G'(w)$ such that  $G'(w) = \bigcup_{z \in T_w}B(z)$ and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{definition}

<!-- \begin{split} -->
<!-- A_{B, M}(z) & = -->
<!-- \begin{cases} -->
<!-- i, & \text{for}\quad z \in T_{i_j} \\ -->
<!-- 0  & \text{otherwise}  -->
<!-- \end{cases} -->
<!-- \end{split} -->

@@

Consider a linear granularity $G'$ such that $G' = Alter^{2}_{(1,-2)}(B, Group_7(B))$, which implies it is made up of shrinking every 1st granule of $Group_7(B)$ by $2$ granules.<!-- $$G'= Alter^{2}_{(1,-1)}(BG, Group_3(B))$$ and --><!--This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different.--> Then $Q_{B, G'}$ is a repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$. Here, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$. Hence using \ref{def:quasicircular}, then we will have:

@@

Consider a school semester which always lasts for 18 weeks and 2 days, starting with one week of orientation followed by an in-session period of 6 weeks, a semester-break of 1 week and again an in-session period of 7 weeks and 1 week study break before final exams which continue for 9 days. Let the linear granularities $M_1$ and $M_2$ denote in-session semester period and semester break periods respectively. Both $M_1$, $M_2$ and $M = M_{1}\bigcup M_{2}$ denoting the semester week type are aperiodic with respect to days ($B$) or weeks ($G$). Hence $A_{B, M}$ denoting day of the semester week type would be an aperiodic cyclic granularity. This is because the placement of the semester within an year would vary across years (\autoref{fig:aperiodic-example} c). \autoref{fig:aperiodic-example} b shows the stack display of time with granules representing same categories (semester break/in-session) stacked on top of one another. \autoref{fig:aperiodic-example} (a) representing the linear display of time is useful in manifesting how the categories would be determined with respect to the bottom granularities using Definition \ref{def:aperiodic}. It is interesting to note here that $Q_{H, M}$ denoting semester week of the semester week type would be a quasi-circular granularity since the distribution of semester weeks within a semester is assumed to remain constant over years. Here $Q_{H, M}$ with a period of 18.3 weeks will have irregular mapping since each semester week types consists of different number of semester weeks.

@@

```{r aperiodic-example, echo=FALSE, out.width = "100%", fig.cap = "Quasi-circular and aperiodic cyclic granularities illustrated through linear (a) and stack (b) display of time. The linear display shows linear granularities days, weeks, semester weeks, semester week type distributed linearly from past to future. Here a semester lasts for 18.3 weeks, each starting with one week of orientation followed by an in-session period of 6 weeks, a semester-break of 1 week and again an in-session period of 7 weeks and 1 week study break before final exams which continue for 1.3 weeks. This pattern remains same for all semester and hence $Q_{H, M}$ with $P = 18.3 weeks$ denoting semester week of the semester week type will be a quasi-circular granularity with repeating patterns. $A_{B, M}$ will be an aperiodic cyclic granularity since the placement of the semester within an year varies across years (shown through c)."}

@@

**Hierarchy table** Let $H_n: (G, C, K)$ be a hierarchy table containing $n$ linear granularities. $G_{l}$ and $G_{m}$ represent the linear granularity of order $l$ and $m$ respectively with $l<m$. $K \equiv P(l,m)$ represents the period length of the grouping $(G_{l}, G_{m})$ for a circular granularity and $K \equiv k(l,m)$ represents the operation to obtain $G_{m}$ from $G_{l}$ for a quasi-circular granularity $C_{G(l),G(m)}$, $\forall l,m \in {1, 2, \dots,n}$. We assume that the cyclic granularity $C_{G(l),G(m)}$ is either circular or quasi-circular. It is difficult to position an aperiodic linear granularity in a hierarchy and hence the possibility of an aperiodic cyclic granularity is not considered.

@@

<!-- The multiple-order-up cyclic granularites resulting from the first two cases are circular and has been handled in \autoref{sec:all-circular}. Any cyclic granularity resulting from the last case are quasi-circular. Examples might include hour-of-month or day-of-year. -->

We consider the case of only one quasi-circular single order-up granularity in the hierarchy table. Any multiple-order-up quasi-circular granularity $C_{l, m}(z)$ can then be obtained as a function of a single (or multiple) order-up circular granularity $C_{l,m'}(z)$ and single order-up quasi-circular granularity $C_{m',m}(z)$ or vice versa. Two different approaches need to be employed leading to the following cases:

<!-- For example, with $C_{l, m}(z) = f(C_{l,m'}(z), C_{m',m}(z))$, -->
- $C_{l,m'}(z)$ is circular and $C_{m',m}(z)$ is quasi-circular
\begin{equation} \label{eq:multifromsingle-quasi1}
\begin{split}
C_{G_l,G_{m}}(z) & = C_{G_{l},G_{m'}}(z) + P(l, m')C_{G_{m'},G_{m}}(z) \\
\end{split}
\end{equation}

@@

<!-- CONTENT CUT -->
<!-- Clearly, the computation of multiple-order-up granularities become more involved with quasi-circular granularities. $C_{hour, month}(z)$ can be represented as function of circular granularity $C_{hour, day}(z)$ and quasi-circular granularity $C_{day, month}(z)$, both of which are single order-up. However, $C_{day, year}(z)$ can not be computed only with single order-up granularities $C_{day, month}(z)$ and $C_{month, year}(z)$. It also requires the knowledge of the composition of the linear granularity $days$ within $months$ in an $year$ for computation of $\vert T_w \vert$. -->
<!-- If $C_{G_{l},G_{m'}}(z)$ or $C_{G_{l},G_{m'}}(z)$ are multiple-order-up circular granularities, they could be computed using similar approach in Section \ref{sec:all-circular}. -->

@@

<!-- CONTENT CHANGE -->
<!-- The representation of single order-up quasi-circular granularities using multiple-order-up quasi-circular granularities is not discussed in this paper and is left for future work. -->

This case is similar to Equation \ref{eq:multifromsingle-quasi2}. Single-order-up quasi-circular granularity ($C_{G_{l},G_{m'}}$) could be obtained from multiple-order-up quasi-circular granularity ($C_{G_l,G_{m}}$) and single/multiple-order-up circular granularity $C_{m',m}$ using \ref{eq:multifromsingle-quasi2}.
